---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(ggplot2)
mtcars$am <-as.
knitr::opts_chunk$set(echo = FALSE)
```


## The `ggplot` approach

### A quick lesson from art

***Have you ever watched Bob Ross paint?***

You should! Here is an example of someone painting in the style of Bob Ross on an iPad:

![(Image courtesy of instagram: [procreatewithbobross](http://www.instagram.com/procreatewithbobross))](images/bob-ross-ex.gif)<br></br>

Notice how the painting comes together in steps, building to a beautiful final product:

1. From a blank canvas, a black foreground and light-blue background with clouds add simple dimension;<br></br>
2. Next, the mountains add significant substance to the painting;<br></br>
3. Embellishments further fill out the scene, like the different color snow on the east and west sides of the mountain, or the small forest at the base;<br></br>
4. Finally, color and detail fill the foreground finishing the work.

### "Painting" a plot
When we use `ggplot2` to make our graphics, we will take the artistic approach, proceeding incrementally.  Starting with a background, we add basic structure, then substance, embellishments, and finish with important details.  We will work sequentially, adding and modifying until our final plot makes us happy, like Bob's trees.

![(Images made with R and compiled into a gif via [ezgif.com](https://ezgif.com/maker))](images/plot-ross-ex.gif)<br></br>

Notice how this plot above is built in steps:

1. From a blank gray canvas, we add variables (e.g. Sepal.Width) to the axes to provide simple dimension;<br></br>
2. Next, points create a scatter plot, demonstrating the relationship between our variables;<br></br>
3. Embellishments add to the plot, like coloring the dots according to the species of iris;<br></br>
4. Finally, we adjust our axes labels and scales, and add titles and themes to finish the plot.
<br></br>

**In the next topic, we'll walk through the incremental steps required to build this very plot!**

## Painting a plot, Bob Ross style

To get our a fresh canvas, we simply start with the command `ggplot()`; try it out for yourself
```{r base-background, exercise=TRUE}
ggplot()
```

### Adding simple dimensions
For our first plot, we will use the `iris` data frame.  It contains measurements from 150 iris flowers, covering three different species: setosa, versicolor, and virginica.  For each flower, we record the species of iris, as well as length and width measurements of the flower's sepals and petals. 

Here is a sneak peak of the data frame.
```{r,echo=F}
head(iris,3)
```

When we know the data frame from which we want to build our plot (and it's in `tidy`-format, more on this later), we can add this to the code for out plot.  

In our painting analogy, this is like putting pain on a pallete. Setting `data = iris` allows us to pull data from the `iris` data frame and use it to build the plot. Let's add some simple dimension to our blank canvas by

1. specifying a data frame (painting pallete)
2. specifying which variables we want to appear on our axes.

In the code below, replace the `...` with two variables of interest; specifically, put `Sepal.Width` on the $y$-axis and `Petal.Length` on the $x$-axis:

```{r ggplot-axes-aes, exercise=TRUE, exercise.lines = 5}
ggplot(data = iris, 
       aes(x = ...,
           y= ...))
```

```{r ggplot-axes-aes-hint}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width))
```

### Adding substance
To highlight the relationship between the variables on our axes, we use a `geometry`.  In the case where we work with two quantitative variables, we often use a "point", or dot, to represent each flower in the data frame.

Let's *add* the point geometry to our plot. Run the code below!
```{r ggplot-geom, exercise=TRUE}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width)) +
    geom_point()
```
Notice we accomplished this by `+` (adding) `geom_point()` (a point geometry) to the canvas we already set up.

> *Important Note:* the plus sign **must be on the same line** as the last element you added to your plot; then hit return to add your next element.

Before continuing to the next topic, please explore this basic "painting" by changing around the variables you use for `x` and `y`. Specifically, what happens when you set one of your axes to the *categorical* variable `Species`. What happens, and why?

#### Exercise:
In the block below, 

* practice writing the additional code to add the point geometry to the plot;
* then, experiment with different variables for the `x` and `y` aesthetics.

```{r ggplot-geom-explore, exercise=TRUE, exercise.lines = 6}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y= Sepal.Width))
```

```{r ggplot-geom-explore-hint-1}
# Change the variables specified in the x and y aesthetics
ggplot(data = iris, 
       aes(x = Petal.Length,
           y= Sepal.Width))+
    geom_point()
```

```{r ggplot-geom-explore-hint-2}
# Change the variables specified in the x and y aesthetics
ggplot(data = iris, 
       aes(x = Species,
           y= Sepal.Width))+
    geom_point()
```

```{r ggplot-geom-explore-hint-3}
# Change the variables specified in the x and y aesthetics
ggplot(data = iris, 
       aes(x = Petal.Length,
           y= Species))+
    geom_point()
```

## Adding some happy little trees
### Plot embellishments
Here is our plot so far
```{r scatter-plot, echo=T, fig.align='center',fig.height=4,fig.width=4}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width))+
    geom_point()
```

Suppose, however, we'd like to color these points according the `Species` of iris. 

> **Important Concept**: Any component of the plot that depends on data *must be assigned to an "aesthetic", so inside `aes()`*.

Below is an example of adding an aesthetic to the point geometry that accomplishes our goal. Since the aesthetic is applied to the point geometry, one approach is to add `aes(color=Species)` inside our point geometry code. Inspect the block below and try running it.
```{r ggplot-color-aes, exercise=TRUE, exercise.lines=5}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width)) +
    geom_point(aes(color=Species))
```

```{r ggplot-color-aes-hint-1}
# we could also add the aesthetic inside the ggplot() component
# but this means all geometries that follow would have this aesthetic.
# This approach makes color=Species a "global" aesthetic
# The code below also generates the same plot:
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width,
           color = Species)) +
    geom_point()
```
**Explore:** Try changing the color aesthetic to depend on a different variable, like `Sepal.Width`.  What happens? Can you explain *why* this happens?

Notice that when adding a color aesthetic to the plot, a legend is automatically created and appended to the side.^[We can change the position of the legend; another time, perhaps.]


## Finishing touches

The `ggplot2` package allows for nearly endless modifications and finishing touches.  Some of these include:

* adjusting the units of our axes and changing which number divisions show
* adding trend lines to scatter plots
* adding text and shape annotations on top of the plot
* applying general [plotting themes](https://ggplot2.tidyverse.org/reference/ggtheme.html) 

In the interests of time and attention span, however, we will consider one more useful step: *adding axis labels to the plot*.

Since we are adding a layer, we put a `+` after our last line of code. Then, after hitting enter to get to the next line, we will use the `labs()` element to change our axes. 

Let's see if your `ggplot2` intuition is developing. Run the code below, and then make the necessary changes provide meaningful labels to the axes and the title.

```{r ggplot-axis-labels, exercise=TRUE, exercise.lines=8}
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width)) +
    geom_point(aes(color=Species)) + 
    labs(x=" some x label",
         y="some y label",
         title="title anyone?")
```

```{r ggplot-axis-labels-hint-1}
# here is one option
ggplot(data = iris, 
       aes(x = Petal.Length,
           y = Sepal.Width)) +
    geom_point(aes(color=Species)) + 
    labs(x="Petal length in inches",
         y="Sepal width in inches",
         title="Comparing sepal width and petal length among species of irises")
```


## Checking your understanding!
In this final section, we review with a couple of questions about what we have learned in this lesson. Feel free to flip back through the exercises and instructions in the previous topics.  In the last block, I challenge you to recreate a new plot based on the `mtcars` data frame!

You may want to try running some R code as you answer questions, so please use the box below to do so!
```{r ggplot-testing, exercise=TRUE, exercise.lines=10}
# Place your test code here


```


### Review Quiz 

**Question 1**
```{r question-1}
  question("Which code below correctly associates the iris data frame to the plot we want to build?",
    answer("ggplot() + iris"),
    answer("ggplot() + data = iris"),
    answer("ggplot(data = iris)", correct = TRUE),
    answer("iris(ggplot())"),
    allow_retry = T,
    random_answer_order = T,
    incorrect = paste(random_encouragement(),"See the first topic: Painting a plot, Bob Ross Style"),
    post_message = random_praise()
  )
```

**Question 2**
```{r question-2}
    question("Assume we're working with the `mtcars` data frame. Which code below correctly puts the `mpg` variable on the x axis and the `hp` variable on the y axis?",
    answer("True", correct = TRUE),
    answer("False1"),
    answer("False2"),
    answer("False3"),
    allow_retry = T,
    random_answer_order = T,
    incorrect = random_encouragement(),
    post_message = random_praise()
  )
```


**Question 3**: Please refer to the code block below to answer this question.

```
ggplot(data = mtcars, aes(x=mpg,y=hp))
+
geom_point()
```

```{r question-3}
question("True or False: the command written in Code Block 1 would result in an error.",
    answer("True", correct = TRUE),
    answer("False"),
    allow_retry = T,
    random_answer_order = T,
    incorrect = paste(random_encouragement(),"See the 'Important Note' in the first topic: Painting a plot."),
    post_message = random_praise()
  )
```

**Question 4**
```{r question-4}
    question("Which of the following statements are true? Select all that apply",
    answer("Plot elements that depend on data are called 'aesthetics' and must appear inside aes()", correct = TRUE),
    answer("A categorical variable, like Species in the iris data frame, cannot appear as the aesthetic for the x axis."),
    answer("The labs() element helps you assign labels to your axes and set a title for the plot.", correct = TRUE),
    answer("The point geometry, geom_point(), is the only geometry that ggplot can use."),
    allow_retry = T,
    random_answer_order = T,
    incorrect = paste(random_encouragement(),"See the 'Important Note' in the first topic: Painting a plot."),
    post_message = random_praise()
  )
```

***

### Final Exercise
Using what you have learned from this tutorial, please use `ggplot2` to recreate the following plot!
