---
title: "Example `learnr` tutorial with CSS"
output: learnr::tutorial
css: css/custom.css
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
knitr::opts_chunk$set(echo = FALSE)
```


## Exploring some html options

Here is some "paragraph" text.

<p id="para1"> Here is text with a `#para1` id </p>

#### A heading 4
with some text.

<button class="exer-button">
    Exercise!
</button>
<div class="exer-block">

```{r test, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r qu1}
question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  )
```

</div>

And we can type outside of the exercise



[drop-down click?]

### Exercise reveal answer {.tabset .tabset-fade}
This is done through R markdown by specifying `{.tabset}` after a section. The subsections that follow end up in a tabset fashion. Could be used for Question/Answer reveals. You can also try `{.tabset-dropdown}`.

#### Coding
```{r ex2, exercise=TRUE, exercise.eval=TRUE}
mtcars
```


#### Quiz

```{r quiz3}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```



## Playing with html stuff

### Exercise 

#### Test
Here is some text

#### Other
Here is some other text


<!-- <div class="exercise"> -->
<!--   <button class="exercise-button">Exercise</button> -->
<!--   <div class="exercise-content"> -->
<!--     Howdy y'all -->
<!--   </div> -->
<!-- </div> -->

<!-- <ul class="specialtabs"> -->
<!--   <li class="active">one</li>  -->
<!--   <li>Two</li>  -->
<!-- </ul> -->
***
***
***
***

## Topic 3

### Making custom questions?
[Learnr Tutorial for custom questions](https://learnr-examples.shinyapps.io/quiz_question/#section-custom-question-types)

```{r custom_class, echo = FALSE}
question_is_correct.always_correct <- function(question, value, ...) {
  learnr::mark_as(TRUE, messages = NULL)
}

registerS3method("question_is_correct", "always_correct", question_is_correct.always_correct)

ques <- question(
  "Custom Method",
  answer("answer", correct = TRUE),
  answer("also marked as correct", correct = FALSE),
  type = c("always_correct", "learnr_checkbox")
)
```

```{r qu-custom}
ques
```


### Exercise with Hint

*Here's an exercise where the chunk is pre-evaluated via the `exercise.eval` option (so the user can see the default output we'd like them to customize). We also add a "hint" to the correct solution via the chunk immediate below labeled `print-limit-hint`.*

Modify the following code to limit the number of rows printed to 5:

```{r print-limit, exercise=TRUE, exercise.eval=TRUE}
mtcars
```

```{r print-limit-hint}
head(mtcars)
```


### Shiny Example
```{r, echo=FALSE}
sliderInput("bins", "Number of bins:", min = 1, max = 50, value = 30)
plotOutput("distPlot")
```

```{r, context="server"}
output$distPlot <- renderPlot({
  x <- faithful[, 2]  # Old Faithful Geyser data
  bins <- seq(min(x), max(x), length.out = input$bins + 1)
  hist(x, breaks = bins, col = 'darkgray', border = 'white')
})
```

### Quiz

*You can include any number of single or multiple choice questions as a quiz. Use the `question` function to define a question and the `quiz` function for grouping multiple questions together.*

Some questions to verify that you understand the purposes of various base and recommended R packages:

```{r quiz}
quiz(
  question("Which package contains functions for installing other R packages?",
    answer("base"),
    answer("tools"),
    answer("utils", correct = TRUE),
    answer("codetools")
  ),
  question("Which of the R packages listed below are used to create plots?",
    answer("lattice", correct = TRUE),
    answer("tools"),
    answer("stats"),
    answer("grid", correct = TRUE)
  )
)
```

## Resources to Explore

* [Translations and controls](https://rstudio.github.io/learnr/articles/multilang.html)

* 
