---
title: "Making Summary Tables in the Tidyverse"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo = FALSE)
titanic<-read.csv('data/titanic2.csv', header = T,
                  stringsAsFactors = T)
titanic<-titanic %>% mutate(
    Fare = round(Fare,2)
)
```

## Categorical data with table() and prop.table()
```{r}
# Just counts
titanic %>% with(
    table(Sex, Survived)
)

titanic %>% with(
    table(Sex, Survived)
) %>% addmargins()

# Row percentages
titanic %>% with(
    table(Sex, Survived)
) %>% prop.table(margin =1)

#column percentages
titanic %>% with(
    table(Sex, Survived)
) %>% prop.table(margin =2)

# total sample
titanic %>% with(
    table(Sex, Survived)
) %>% prop.table()
```


## Quantitative data
```{r}
titanic %>% summarise(meanFare = mean(Fare),
                      sdFare = sd(Fare),
                      IQR_age = IQR(Age))

titanic %>% summarise(meanFare = mean(Fare),
                      sdFare = sd(Fare),
                      IQR_age = IQR(Age))

titanic %>% summarise(meanFare = mean(Fare),
                      sdFare = sd(Fare),
                      IQR_age = IQR(Age)) %>%
    pivot_longer(cols=everything())
```


## Categorical and Quantitative data, group_by() and summarise()
```{r}
titanic %>% group_by(Pclass) %>%
    summarize(meanAge = mean(Age),
              meanFare = mean(Fare),
              perc_total = n()/nrow(titanic),
              Percent_Survived = sum(Survived=="Yes")/n(),
              sample_size = n())

titanic %>% group_by(Pclass) %>%
    summarize(meanAge = mean(Age),
              meanFare = mean(Fare),
              perc_total = n()/nrow(titanic),
              Percent_Survived = sum(Survived=="Yes")/n(),
              sample_size = n()) %>%
    pivot_longer(cols = -Pclass, names_to = "Summary") %>%
    mutate(Pclass = paste("Passenger Class = ",Pclass)) %>%
    pivot_wider(names_from = Pclass,
                values_from = value) %>%
    kbl() %>% kable_styling("striped")
```
Using "Export" can save as image, or copy to clipboard and paste into a google doc.

*Rounding can be a pain in the ass this way.*
Fix:

* Before switching columns and rows, round each value and then change it to a character, which automatically drops the trailing zeros.
* Use `t()` to switch columns and rows
* Send to `kbl()` to create the table.
```{r}
titanic %>% group_by(Pclass) %>%
    summarize(`Average Age` = mean(Age) %>% round(1) %>%as.character(),
              `Average Fare` = mean(Fare) %>% round(2)%>%as.character(),
              `Percent of all aboard` = (n()/nrow(titanic)) %>%round(3) %>% as.character(),
              `Survival Rate` = (sum(Survived=="Yes")/n()) %>%round(3) %>% as.character(),
              `Total` = as.integer(n()) %>%as.character()) %>%
    mutate(Pclass = paste("Passenger Class",Pclass)) %>%
    remove_rownames() %>% column_to_rownames(var="Pclass") %>%
    t() %>% kbl() %>% kable_styling("striped")
```


